OneD: increasing reproducibility of Hi-C Samples with abnormal karyotypes

### 2017-06-07 main.tex

\author[Vidal \textit{et~al}.]{Enrique Vidal\,$^{\text{\sfb 1,2}*}$,
François le Dily\,$^{\text{\sfb 1,2}}$, Javier Quilez\,$^{\text{\sfb
1,2}}$, Ralph Stadhouders\,$^{\text{\sfb 1,2}}$, Yasmina
Cuartero\,$^{\text{\sfb 1,2}}$, Miguel Beato\,$^{\text{\sfb 1,2}}$ and
Guillaume Fillion\,$^{\text{\sfb 1,2}}$}

@@

\abstract{
  \textbf{Motivation:} The three-dimensional conformation of genomes is an
essential component of their biological activity. The advent of the Hi-C
technology enabled an unprecedented advance in our understanding of genome
structures. However, Hi-C is subject to systematic biases that can
compromise downstream analyses. Several strategies have been proposed to
remove those biases, but none of them addresses the common issue of
abnormal karyotypes. Many experiments are performed in cancer cell lines,
which typically harbor large-scale copy number variations that create
visible defects on the raw Hi-C maps. The consequences of these widespread
artifacts on the normalized maps are mostly unexplored.\\
	\textbf{Results:}
We observed that current normalization methods perform badly in the
presence of large-scale copy number variations, obscuring biological
variations and enhancing batch effects. To address this issue, we
developed an alternative approach designed to take into account chromosomal
abnormalities. The method, called OneD, increases reproducibility among
replicates of Hi-C samples with abnormal karyotype, thus significantly
outperforming previous methods.  In the absence of chromosomal
aberrations, OneD fared equally well as state-of-the-art methods, making
it a safe choice for Hi-C normalization. OneD is fast and scales well in
terms of computing resources for resolutions up to 1 kbp.\\	
\textbf{Availability:} OneD is implemented as an R package available at
\href{http://www.github.com/qenvio/dryhic}
{http://www.github.com/qenvio/dryhic}.\\	
\textbf{Contact:} \href{enrique.vidal@crg.eu}{enrique.vidal@crg.eu}\\	
\textbf{Supplementary information:} Supplementary data are available at
\textit{Bioinformatics}	online.}

@@

One of the crown achievements of modern biology was to realize that
genomes have an underlying three-dimensional structure contributing to
their activity \citep{rowley2016three}. In mammals, this organization
plays a key role in guiding enhancer-promoter contacts (CITE), in V(D)J
recombination (CITE) and in X chromosome inactivation (CITE). One of the
most significant breakthroughs towards this insight was the development of
the high throughput chromosomal conformation capture technology (Hi-C),
assaying chromosomal contacts at a genome-wide scale
\citep{lieberman2009comprehensive}. Nowadays, exploring the spatial
organization of chromatin has become a priority in many fields and
Hi-C has become part of the standard molecular biology toolbox (CITE).

Contrary to the precursor technologies 3C, 4C and 5C \citep{de2012decade},
Hi-C interrogates all possible pairwise interactions between restriction
fragments. However, this does not guarantee that the method has no bias.
On the contrary, local genome features such as the G+C content, the
availability of restriction enzyme sites and the mappability of the
sequencing reads have been shown to impact the results
\citep{yaffe2011probabilistic}. In addition, the general experimental
biases such as batch effects and protocol variation also apply. It is thus
important to normalize Hi-C data in order to remove biases and artifacts,
so that they are not confused with biological signal.

Several methods have been proposed to remove biases in Hi-C experiments
\citep{schmitt2016genome}. The first strategy is to model biases
explicitly from a defined set of local genomic features, such as the G+C
content and the availability of restriction sites. This approach is used
in the method of \cite{yaffe2011probabilistic} and in Hicnorm by
\cite{hu2012hicnorm}. The second strategy is to implicitly correct
unknown biases by enforcing some regularity on the data. This approach is
used in the ICE method of \cite{imakaev2012iterative}, where the total
amount of contacts of every bin is imposed to be the same. ICE is
currently the most popular method, due in part to its speed.

Neither of these strategies were designed with regard for cell types with
karyotypic aberrations, most common in cancer cells in culture.  Hi-C is
very sensitive to aneuploidy, copy number variations and translocations.
Actually, these aberrations have so much influence on the outcome that the
artifacts can be used to re-assemble the target genome
\citep{korbel2013genome}.  So far the only attempt to address the issue
was the correction method caICB by \cite{wu2016computational}. However,
caICB applies a chromosome-wide correction, effectively excluding the
numerous cases of partial aneuploidy and regional copy number variations.

Here we propose \textit{OneD}, a method to correct local chromosomal
abnormalities. OneD explicitly models the contribution of known biases via
a generalized additive model. The normalized data is more reproducible
between replicates and across different protocols. Importantly, OneD is
also applicable when cells have a normal karyotype, where it performs as
well as the best normalization methods. Finally, the implementation is
faster than ICE and scales up to 1 kbp resolution with reasonable
computing resources.

@@

Our approach is to model the tally of contacts for each bin, thus reducing
the matrix to a one dimension score (hence the name OneD). We assume that
the total number of contacts per bin ($t_{i}$) can be approximated by a
negative binomial distribution. This choice is sensible because the amount
of contacts is a discrete variable and because the negative binomial
distribution allows for overdispersion. We further assume that the
explicit sources of bias have independent contributions to the mean of the
distribution for a given bin ($\lambda_i$).

@@

\begin{table*}
\processtable{Hi-C data sets used in this study.
\label{tab:samples}}
{\begin{tabular}{ccccccc}
  \toprule
  \textbf{Sample ID} & \textbf{Cell type} & \textbf{Application} &
  \textbf{RE} & \textbf{Sequencing core} & \textbf{Set(s)} &
  \textbf{Source} \\
  \midrule
dc3a1e069\_51720e9cf & T47D & \textit{in situ} Hi-C &
  DpnII & CRG & T47D, K562 & NA \\
b1913e6c1\_51720e9cf & T47D & \textit{in situ} Hi-C &
  DpnII & CRG & T47D & NA \\
dc3a1e069\_ec92aa0bb & T47D & \textit{in situ} Hi-C &
  DpnII & CRG & T47D, K562 & NA \\
HindIII\_T0 & T47D & dilution Hi-C &
  HindIII & CRG & T47D & SRR1054341 \\
NcoI\_T0    & T47D & dilution Hi-C &
  NcoI      & CRG & T47D & SRR1054343 \\
ENCLB758KFU & T47D & dilution Hi-C &
  HindIII   & UMass & T47D & ENCLB758KFU \\
ENCLB183QHG & T47D & dilution Hi-C &
  HindIII   & UMass & T47D & ENCLB183QHG \\
HIC069  & K562 & \textit{in situ} Hi-C &
  MboI & Baylor & T47D, K562 &    SRR1658693 \\
HIC070  & K562 & \textit{in situ} Hi-C &
  MboI & Baylor & T47D, K562 &    SRR1658694 \\
HIC071  & K562 & \textit{in situ} Hi-C &
  MboI & Baylor & K562 & SRR1658695,SRR1658696 \\
HIC074  & K562 & \textit{in situ} Hi-C &
  MboI & Baylor & K562 & SRR1658701,SRR1658702 \\
b7fa2d8db\_bfac48760 & B-cell  & \textit{in situ} Hi-C &
  DpnII & CRG & mm10 & GSE96611 \\
fc3e8b36a\_7bf1bf374 & ES-cell & \textit{in situ} Hi-C &
  DpnII & CRG & mm10 & GSE96611 \\
b7fa2d8db\_7284b867a & B-cell  & \textit{in situ} Hi-C &
  DpnII & CRG & mm10 & GSE96611 \\
fc3e8b36a\_38bfd1b33 & ES-cell & \textit{in situ} Hi-C &
  DpnII & CRG & mm10 & GSE96611 \\
b7fa2d8db\_58e812fc2 & B-cell  & \textit{in situ} Hi-C &
  DpnII & CRG & mm10 & GSE96611 \\
fc3e8b36a\_c990a254e & ES-cell & \textit{in situ} Hi-C &
  DpnII & CRG & mm10 & GSE96611 \\
b7fa2d8db\_73f11d923 & B-cell  & \textit{in situ} Hi-C &
  DpnII & CRG & mm10 & GSE96611 \\
fc3e8b36a\_4bf044f18 & ES-cell & \textit{in situ} Hi-C &
  DpnII & CRG & mm10 & GSE96611 \\
GSM987817 & B-cell  & dilution Hi-C &
  HindIII & UCSC & mm10 & SRR543428-SRR543431 \\
GSM987818 & B-cell  & dilution Hi-C &
  HindIII & UCSC & mm10 & SRR543432-SRR543442 \\
GSM862720 & ES-cell & dilution Hi-C &
  HindIII & UCSC & mm10 & SRR443883-SRR443885 \\
GSM862721 & ES-cell & dilution Hi-C &
  HindIII & UCSC & mm10 & SRR400251-SRR400256 \\
GSM862722 & ES-cell & dilution Hi-C &
  NcoI    & UCSC & mm10 & SRR443886-SRR443888 \\
  \botrule
\end{tabular}}{}
\end{table*}

\subsection{Data processing}

@@

We developed the routines contained in the \texttt{dryhic} R package to
efficiently create sparse representations of contact matrices and further
apply \textit{vanilla}, \textit{ice} and \textit{oned} corrections.
\texttt{HiTC} \citep{servant2012hitc} and \texttt{HiCapp}
\citep{wu2016computational} were used to carry out the \textit{lgf} and
\textit{caib} corrections respectively. All the results are based on a
resolution of 100 kbp, but we found no major differences for different
values (not shown).

@@

We benchmarked \textbf{\textit{oned}}, against \textbf{\textit{vanilla}},
\textbf{\textit{ice}} \citep{imakaev2012iterative}, \textbf{\textit{caib}}
\citep{wu2016computational} and \textbf{\textit{lgf}}
\citep{hu2012hicnorm, servant2012hitc}. The first three methods correct
biases implicitly, whereas the fourth method does it explicitly.

To measure the gain or loss of similarity upon normalization, we compared
raw matrices to obtain a baseline. The differences with this baseline were
estimated using a linear mixed model fitted with the \texttt{lmer}
function of the \texttt{lme4} R package \citep{bates2015lme4}, where the
fixed effect was the normalization method and the random effect was the
chromosome. Receiver operating characteristic (ROC) curves were then
computed using \texttt{ROCR} package \citep{sing2005rocr}.

@@

\subsection{Correction method}

The principle of OneD is to explicitly model Hi-C biases on a single
variable: the total amount of contacts for each bin of the matrix (see
\ref{sec:model}). The reason for this choice is that the total amount of
contacts is approximately proportional to the local copy number. For
instance, a duplicated region in a diploid genome will show on average a
50\% increase in the number of contacts. Discontinuities of the amount of
contacts thus correspond to changes of the copy number.

\begin{figure}[!tp]
\centerline{\includegraphics[width=.45\textwidth]{img/figure1.pdf}}
\caption{A. Non linear relationship between the number of restriction
sites per bin and the total number of contacts per bin in T47D. B. Total
number of contacts per bin in chromosome 17 of T47D. Raw (brown) and
corrected (blue) signals are shown. The region that corresponds to the
first 20 Mbp of the chromosome is present in single copy in this clone of
T47D, explaining that the signal is approximately half compared to the
rest of the chromosome.  Also notice that the raw signal is noisier than
the corrected one.}
\label{fig:totals}
\end{figure}

Other biases affect the total amount of contacts in a continuous but not
necessarily linear way. Figure~\ref{fig:totals}A shows the relationship
between the amount of contacts and the number of restriction enzyme sites
in T47D, a breast cancer cell line with aberrant and unstable karyotype.
Four clouds are visible. Each corresponds to a copy number between one and
four. In all of them, the relationship flattens as the number of
restriction sites increases. For this reason, OneD fits a non-linear
relationship between the total amount of contacts and the known sources of
bias (by default the G+C content, the number of restriction sites and the
mappability of the reads).

The biases are estimated genome-wide and each cell of the matrix is then
corrected using equation (\ref{eq:xhat}). Note that the corrected amount
of contacts is still proportional to the copy number. The correction thus
represents a more stable and precise estimation of the copy number.
Figure~\ref{fig:totals}B shows the corrected number of contacts along
chromosome 17 of a T47D. OneD greatly reduces the wiggling of the total
amount of contacts.

We tested the validity of this approach against the Catalogue Of Somatic
Mutations In Cancer \citep[COSMIC,][]{forbes2010cosmic}. Figure
\ref{fig:copy_number} shows the Pearson correlation between the corrected
number of contacts and the copy number estimation for both T47D and K562
cell lines. Similar results were obtained using Spearman correlation
(Supplementary Figure 2). All the methods except OneD decrease the
agreement with the copy number because they partially correct this bias.
In contrast, OneD enhances the conformity of the signal with the copy
number. Not correcting for variable copy number at that stage may seem to
defeat the purpose of a normalization method, but we will see below how
this can lead to better performance.

\begin{figure}
	\centerline{\includegraphics[width=.45\textwidth]
{img/copy_number_figure2.pdf}}
	\caption{
    Pearson correlation between total number of contacts per bin and
independent copy number estimation (COSMIC) for each of the methods
compared. Left panel T47D breast cancer cell line, right panel K562
leukemia cell line. The new proposal (in blue) outperforms the rest of
alternatives.  }\label{fig:copy_number}
\end{figure}

@@

We first benchmarked the performance of OneD on biological samples with an
aberrant karyotype. A good normalization method should increase the
similarity between biological replicates by reducing irrelevant
experimental variation, such as batch effects, laboratory of origin and
protocol variations. Similarly, a good normalization should decrease the
similarity between samples that received a different treatment in order to
enhance the biological differences.

We assembled two Hi-C data sets obtained from T47D and K562, two cancer
cell lines with aberrant karyotypes. In each set we spiked two
samples from the other cell line (see Table~\ref{tab:samples}) in order to
introduce biological variability. We measured the Pearson correlation of
observed over expected counts to compare matrices before and after
normalization by different methods (see \ref{sec:comp}). This gave an
indication of the impact of a given normalization method. The results are
summarized in Figure~\ref{fig:aberrant}.

The \textit{caib} and \textit{ice} methods increased the similarity
between the different cell lines (Figures~\ref{fig:aberrant}A and
\ref{fig:aberrant}B and Supplementary Figure~3). This is an undesirable
effect, as it obscures the biological variability. In the same vein, these
methods decreased the similarity between samples that received the same
treatment (Figure~\ref{fig:aberrant}A), suggesting that the normalization
process is detrimental to the biological signal in these two cases. The
method \textit{vanilla} followed the same trend but to a lesser extent,
consistent with the fact that it consists of a single \textit{ice}
iteration.

OneD was the only method to increase the similarity between experiments
carried out on the same material but with a different protocol
(Figure~\ref{fig:aberrant}A). In these conditions, \textit{lgf} was
decreasing the similarity. Taken together, the results show that OneD
enhanced biological variation while decreasing experimental variation.

@@

All the methods, including the absence of normalization, succeeded in
recognizing the T47D outliers among the K562 samples, but recognizing the
K562 outliers among the T47D samples proved more challenging. OneD
increased the discrimination power compared to the raw matrices, but all
the other methods decreased it. Actually, they performed little better
than random on this task.

@@

Taken together, these results show that OneD enhances biological variation
and reduces experimental noise on samples with an aberrant karyotype.

\begin{figure}
\centerline{\includegraphics[width=.50\textwidth]{img/correlation_aberrant_figure3.pdf}}
\caption{
Results of the comparison between samples with aberrant karyotype. A and
B. Average changes compared to raw on the T47D and K562 sets. The bars
represent 95\% confidence intervals centered on the mean difference of the
correlation score between a given correction method and the raw data. The
brown dashed line indicates the value of the average score on raw matrices
(set to 0). C and D. ROC curves on the T47D and K562 sets. The areas under
the curve are indicated in the bottom right corner. The color code is the
same as panels A and B. The brown lines correspond to raw matrices. All
results in this figure are based on Pearson correlations between the
observed over expected counts.}
\label{fig:aberrant}
\end{figure}

@@

Does the performance of OneD with aberrant karyotypes comes at the cost of
decreased performance with normal karyotypes? To address this question, we
assembled another data set comprised of normal mouse B cells and embryonic
stem (ES) cells. The cell types were pooled in almost equal proportions
(see Table~\ref{tab:samples}) and the same tests as above were performed.

In these conditions, variations of the protocol had a strong effect on the
impact of the different normalization methods (Figure~\ref{fig:normal}A).
For instance, \textit{caib} and \textit{ice} increased the similarity when
the protocols were different, but decreased it when the protocols were the
same. The effect was stronger when comparing identical cell types, but the
same trend appeared when comparing different cell types, indicating that
these methods may enhance or reduce biological variation, depending on the
context. Once more, \textit{vanilla} followed the same trend as
\textit{ice} to a lesser extent. The \textit{lgf} method increased the
similarity when comparing the same cells with different experimental
protocols, and had little to no effect in the other cases. This indicates
that \textit{lgf} is a safe choice in this case.

OneD decreased the similarity between different cell types when using the
same protocol and increased it between identical cell types when using
different protocols. In this other two cases, it had little effect. In
summary, OneD never enhanced the experimental noise and it reduced it in
one more case than \textit{lgf}.

When interpreting the similarity scores as classification scores, we
observed that all the methods could identify approximately 50\% of the
 biological pairs, after which their performance diverged
(Figure~\ref{fig:normal}B). OneD achieved the highest area under the curve
on this problem, even though it was not always above the others methods at
every position of the graph.

Using other metrics to compare matrices gave similar results
(Supplementary Figures 6 and 7). OneD was always among the top scoring
methods. In these conditions, Spearman correlation of contacts again
appeared as as the worst comparison metric as it showed a lower
performance for all the normalization methods.

Taken together, these results indicate that OneD performs as well as the
best normalization methods even with normal karyotypes.

>>>>

\author[Vidal \textit{et~al}.]{Enrique Vidal\,$^{\text{\sfb 1,2}*}$,
François le Dily\,$^{\text{\sfb 1,2}}$, Javier Quilez\,$^{\text{\sfb
1,2}}$, Ralph Stadhouders\,$^{\text{\sfb 1,2}}$, Yasmina
Cuartero\,$^{\text{\sfb 1,2}}$, Thomas Graf\,$^{\text{\sfb 1,2}}$, Miguel
Beato\,$^{\text{\sfb 1,2}}$ and Guillaume J. Filion\,$^{\text{\sfb 1,2}}$}

@@

\abstract{
  \textbf{Motivation:} The three-dimensional conformation of genomes is an
essential component of their biological activity. The advent of the Hi-C
technology enabled an unprecedented progress in our understanding of
genome structures. However, Hi-C is subject to systematic biases that can
compromise downstream analyses. Several strategies have been proposed to
remove those biases, but the issue of abnormal karyotypes received little
attention. Many experiments are performed in cancer cell lines, which
typically harbor large-scale copy number variations that create visible
defects on the raw Hi-C maps. The consequences of these widespread
artifacts on the normalized maps are mostly unexplored. \\
	\textbf{Results:}
We observed that current normalization methods are not robust to the
presence of large-scale copy number variations, potentially obscuring
biological variations and enhancing batch effects. To address this issue,
we developed an alternative approach designed to take into account
chromosomal abnormalities. The method, called \textit{OneD}, increases
reproducibility among replicates of Hi-C samples with abnormal karyotype,
thus significantly outperforming previous methods.  In the absence of
chromosomal aberrations, \textit{OneD} fared equally well as
state-of-the-art methods, making it a safe choice for Hi-C normalization.
\textit{OneD} is fast and scales well in terms of computing resources for
resolutions up to 1 kbp. \\
\textbf{Availability:} \textit{OneD} is implemented as an R package
available at \href{http://www.github.com/qenvio/dryhic}
{http://www.github.com/qenvio/dryhic}. \\
\textbf{Contact:} \href{enrique.vidal@crg.eu}{enrique.vidal@crg.eu}\\	
\textbf{Supplementary information:} Supplementary data are available at
\textit{Bioinformatics}	online.}

@@

One of the crown achievements of modern biology was to realize that
genomes have an underlying three-dimensional structure contributing to
their activity \citep{rowley2016three}. In mammals, this organization
plays a key role in guiding enhancer-promoter contacts
\citep{de2013topology}, in V(D)J recombination \citep{choi2014ctcf} and in
X chromosome inactivation \citep{galupa2015x}.

One of the most significant breakthroughs towards this insight was the
development of the high throughput chromosomal conformation capture
technology (Hi-C), assaying chromosomal contacts at a genome-wide scale
\citep{lieberman2009comprehensive}. Nowadays, exploring the spatial
organization of chromatin has become a priority in many fields and Hi-C
has become part of the standard molecular biology toolbox
\citep{denker2016second}.

Contrary to the precursor technologies 3C, 4C and 5C \citep{de2012decade},
Hi-C interrogates all possible pairwise interactions between restriction
fragments. However, this does not guarantee that the method has no bias.
On the contrary, local genome features such as the G+C content, the
availability of restriction enzyme sites and the mappability of the
sequencing reads have been shown to impact the results
\citep{yaffe2011probabilistic}. In addition, general experimental biases
such as batch effects. It is thus important to normalize Hi-C data in
order to remove biases and artifacts, so that they are not confused with
biological signal.

Several methods have been proposed to remove biases in Hi-C experiments
\citep{schmitt2016genome}. The first strategy is to model biases
explicitly from a defined set of local genomic features, such as the G+C
content. This approach is used in the method of
\cite{yaffe2011probabilistic} and in Hicnorm by \cite{hu2012hicnorm}. The
second strategy is to implicitly correct unknown biases by enforcing some
regularity condition on the data. This approach is used in the
Iterative Correction and Eigenvector decomposition method (\textit{ICE})
of \cite{imakaev2012iterative}, whereby the total amount of contacts of
every bin is imposed to be the same. \textit{ICE} is currently the most
popular method, due in part to its speed.

Neither of these strategies were designed with regard for cell types with
karyotypic aberrations, most common in cancer cells. Hi-C is very
sensitive to aneuploidy, copy number variations and translocations.
Actually, these aberrations have so much influence on the outcome that the
artifacts can be used to re-assemble the target genome
\citep{korbel2013genome}. So far the only attempt to address the issue
was the chromosome-adjusted Iterative Correction Bias method
(\textit{caICB}) of \cite{wu2016computational}.  However, caICB applies a
uniform chromosome-wide correction, effectively excluding the numerous
cases of partial aneuploidy and regional copy number variations.

Here we propose \textit{OneD}, a method to correct local chromosomal
abnormalities in Hi-C experiments. \textit{OneD} explicitly models the
contribution of known biases via a generalized additive model. The
normalized data is more reproducible between replicates and across
different protocols. Importantly, \textit{OneD} is also efficient when
cells have a normal karyotype, where it performs as well as the best
normalization methods. Finally, the implementation is as fast as
\textit{ICE} and it scales up to 1~kbp resolution with reasonable
computing resources.

@@

Our approach is to model the tally of contacts for each bin, thus reducing
the matrix to a one dimension score (hence the name \textit{OneD}). We assume that
the total number of contacts per bin ($t_{i}$) can be approximated by a
negative binomial distribution. This choice is sensible because the amount
of contacts is a discrete variable and because the negative binomial
distribution allows for overdispersion. We further assume that the
explicit sources of bias have independent contributions to the mean of the
distribution for a given bin ($\lambda_i$).

@@

\begin{table*}
\processtable{Hi-C data sets used in this study. Quique you're here.
\label{tab:samples}}
{\begin{tabular}{ccccccc}
  \toprule
  \textbf{Sample ID} & \textbf{Cell type} & \textbf{Application} &
  \textbf{RE} & \textbf{Sequencing core} & \textbf{Set(s)} &
  \textbf{Source} \\
  \midrule
dc3a1e069\_51720e9cf & T47D & \textit{in situ} Hi-C &
  DpnII & CRG & T47D, K562 & NA \\
b1913e6c1\_51720e9cf & T47D & \textit{in situ} Hi-C &
  DpnII & CRG & T47D & NA \\
dc3a1e069\_ec92aa0bb & T47D & \textit{in situ} Hi-C &
  DpnII & CRG & T47D, K562 & NA \\
HindIII\_T0 & T47D & dilution Hi-C &
  HindIII & CRG & T47D & SRR1054341 \\
NcoI\_T0    & T47D & dilution Hi-C &
  NcoI      & CRG & T47D & SRR1054343 \\
ENCLB758KFU & T47D & dilution Hi-C &
  HindIII   & UMass & T47D & ENCLB758KFU \\
ENCLB183QHG & T47D & dilution Hi-C &
  HindIII   & UMass & T47D & ENCLB183QHG \\
HIC069  & K562 & \textit{in situ} Hi-C &
  MboI & Baylor & T47D, K562 &    SRR1658693 \\
HIC070  & K562 & \textit{in situ} Hi-C &
  MboI & Baylor & T47D, K562 &    SRR1658694 \\
HIC071  & K562 & \textit{in situ} Hi-C &
  MboI & Baylor & K562 & SRR1658695,SRR1658696 \\
HIC074  & K562 & \textit{in situ} Hi-C &
  MboI & Baylor & K562 & SRR1658701,SRR1658702 \\
b7fa2d8db\_bfac48760 & B-cell  & \textit{in situ} Hi-C &
  DpnII & CRG & mm10 & GSE96611 \\
fc3e8b36a\_7bf1bf374 & ES-cell & \textit{in situ} Hi-C &
  DpnII & CRG & mm10 & GSE96611 \\
b7fa2d8db\_7284b867a & B-cell  & \textit{in situ} Hi-C &
  DpnII & CRG & mm10 & GSE96611 \\
fc3e8b36a\_38bfd1b33 & ES-cell & \textit{in situ} Hi-C &
  DpnII & CRG & mm10 & GSE96611 \\
b7fa2d8db\_58e812fc2 & B-cell  & \textit{in situ} Hi-C &
  DpnII & CRG & mm10 & GSE96611 \\
fc3e8b36a\_c990a254e & ES-cell & \textit{in situ} Hi-C &
  DpnII & CRG & mm10 & GSE96611 \\
b7fa2d8db\_73f11d923 & B-cell  & \textit{in situ} Hi-C &
  DpnII & CRG & mm10 & GSE96611 \\
fc3e8b36a\_4bf044f18 & ES-cell & \textit{in situ} Hi-C &
  DpnII & CRG & mm10 & GSE96611 \\
GSM987817 & B-cell  & dilution Hi-C &
  HindIII & UCSC & mm10 & SRR543428-SRR543431 \\
GSM987818 & B-cell  & dilution Hi-C &
  HindIII & UCSC & mm10 & SRR543432-SRR543442 \\
GSM862720 & ES-cell & dilution Hi-C &
  HindIII & UCSC & mm10 & SRR443883-SRR443885 \\
GSM862721 & ES-cell & dilution Hi-C &
  HindIII & UCSC & mm10 & SRR400251-SRR400256 \\
GSM862722 & ES-cell & dilution Hi-C &
  NcoI    & UCSC & mm10 & SRR443886-SRR443888 \\
  \botrule
\end{tabular}}{}
\end{table*}

%\subsection{Data processing}

@@

We developed the routines contained in the \texttt{dryhic} R package to
efficiently create sparse representations of contact matrices and further
apply \textit{vanilla}, \textit{ICE} and \textit{oneD} corrections.
\texttt{HiTC} \citep{servant2012hitc} and \texttt{HiCapp}
\citep{wu2016computational} were used to carry out the \textit{LGF} and
\textit{caICB} corrections respectively. All the results are based on a
resolution of 100 kbp, but we found no major differences for different
values (not shown).

@@

To measure the gain or loss of similarity upon normalization, we compared
raw matrices to obtain a baseline. The differences with this baseline were
estimated using a linear mixed model fitted with the \texttt{lmer}
function of the \texttt{lme4} R package \citep{bates2015lme4}, where the
fixed effect was the normalization method and the random effect was the
%chromosome. Receiver operating characteristic (ROC) curves were then
computed using \texttt{ROCR} package \citep{sing2005rocr}.

@@

\subsection{Experimental bias correction}

The principle of \textit{OneD} is to explicitly model Hi-C biases on a
single variable: the total amount of contacts for each bin of the matrix
(see \ref{sec:model} for detail). The reason for this choice is that the
total amount of contacts is approximately proportional to the local copy
number. For instance, a duplicated region in a diploid genome will show on
average a 50\% increase in the number of contacts. Discontinuities of the
amount of contacts thus correspond to changes of the copy number.

\begin{figure}
\centerline{\includegraphics[width=.45\textwidth]{img/figure1.pdf}}
\caption{A. Non linear relationship between the number of restriction
sites per bin and the total number of contacts per bin in T47D. B. Total
number of contacts per bin in chromosome 17 of T47D. Raw (brown) and
corrected (blue) signals are shown. The region that corresponds to the
first 20 Mbp of the chromosome is present in single copy in this clone of
T47D, explaining that the signal is approximately half compared to the
rest of the chromosome.  Also notice that the raw signal is noisier than
the corrected one.}
\label{fig:totals}
\end{figure}

Experimental biases affect the total amount of contacts in a continuous
but not necessarily linear way. Figure~\ref{fig:totals}A shows the
relationship between the amount of contacts and the number of restriction
enzyme sites in T47D, a breast cancer cell line with aberrant and unstable
karyotype.  Four clouds are visible. Each corresponds to sequences present
in one to four copies. In all of them, the relationship flattens as the
number of restriction sites increases. In order to capture this
relationship, \textit{OneD} fits a non-linear model between the total
amount of contacts and the known sources of bias (by default the G+C
content, the number of restriction sites and the mappability of the
reads).

The experimental biases are estimated genome-wide and each cell of the
matrix is then corrected using equation (\ref{eq:xhat}). Note that the
corrected amount of contacts is still proportional to the copy number.
Figure~\ref{fig:totals}B shows the corrected number of contacts along
chromosome 17 of T47D. \textit{OneD} greatly reduces the wiggling of the
total amount of contacts.

We tested the validity of this approach against the Catalogue Of Somatic
Mutations In Cancer \citep[COSMIC,][]{forbes2010cosmic}. In what follows,
we benchmarked \textit{OneD}, against \textit{vanilla}, \textit{ICE}
\citep{imakaev2012iterative}, \textit{caICB} \citep{wu2016computational}
and the Local Genomic Features method
\citep[\textit{LGF},][]{hu2012hicnorm, servant2012hitc}. The first three
methods correct biases implicitly, whereas the fourth method does it
explicitly. Figure \ref{fig:copy_number} shows the Pearson correlation
between the corrected number of contacts and the copy number estimation
for both T47D and K562 cell lines. Similar results were obtained using
Spearman correlation (Supplementary Figure 2). All the methods except
\textit{OneD} decrease the agreement between the signal and the copy
number because they partially correct the discrepancy. In contrast,
\textit{OneD} enhances the conformity of the signal with the copy number.
Not correcting for variable copy number at that stage may seem
counter-intuitive, but the tests below will show this leads to better
performance.

\begin{figure}[b]
	\centerline{\includegraphics[width=.45\textwidth]
{img/copy_number_figure2.pdf}}
	\caption{
Pearson correlation between total number of contacts per bin and
independent copy number estimation (COSMIC) for each of the methods
compared. Left panel T47D breast cancer cell line, right panel K562
leukemia cell line. The new proposal (in blue) outperforms the rest of
alternatives.}
\label{fig:copy_number}
\end{figure}

@@

We first benchmarked the performance of our approach on biological
samples with an aberrant karyotype. A good normalization method should
increase the similarity between biological replicates by reducing
irrelevant experimental variation, such as batch effects, laboratory of
origin and protocol variations. Similarly, a good normalization should
decrease the similarity between samples that received a different
treatment in order to enhance the biological differences.

We assembled two Hi-C data sets obtained from T47D and K562, two cancer
cell lines with aberrant karyotypes. In each set we spiked two samples
from the other cell line (see Table~\ref{tab:samples}) in order to
introduce biological variability. We compared matrices before and after
normalization by different methods using the Pearson correlation of
observed over expected counts (see \ref{sec:comp}). This gave an
indication of the impact of a given normalization method. The results are
summarized in Figure~\ref{fig:aberrant}.

The \textit{caICB} and \textit{ICE} methods increased the similarity
between the different cell lines (Figures~\ref{fig:aberrant}A and
\ref{fig:aberrant}B and Supplementary Figure~3). This is an undesirable
effect, as it obscures the biological variability. In the same vein, these
methods decreased the similarity between samples that received the same
treatment (Figure~\ref{fig:aberrant}A), suggesting that the normalization
process is detrimental to the biological signal in these two cases. The
method \textit{vanilla} followed the same trend but to a lesser extent,
consistent with the fact that it consists of a single \textit{ICE}
iteration. \textit{OneD} was the only method to increase the similarity
between experiments carried out on the same material but with a different
protocol (Figure~\ref{fig:aberrant}A).

@@

All the methods, including the absence of normalization, succeeded in
identifying the T47D outliers among the K562 samples, but recognizing the
K562 outliers among the T47D samples proved more challenging.
\textit{OneD} increased the discrimination power compared to the raw
matrices, but all the other methods decreased it. Actually, they performed
little better than random on this task.

@@

Taken together, these results show that correcting experimental biases
with \textit{OneD} enhances the biological variation and reduces the noise
on samples with an aberrant karyotype.

\begin{figure}
\centerline{\includegraphics[width=.50\textwidth]
  {img/correlation_aberrant_figure3.pdf}}
\caption{
Results of the comparison between samples with aberrant karyotype. A and
B. Average changes compared to raw on the T47D and K562 sets. The bars
represent 95\% confidence intervals centered on the mean difference of the
correlation score between a given correction method and the raw data. The
brown dashed line indicates the value of the average score on raw matrices
(set to 0). C and D. ROC curves on the T47D and K562 sets. The areas under
the curve are indicated in the bottom right corner. The color code is the
same as panels A and B. The brown lines correspond to raw matrices. All
results in this figure are based on Pearson correlations between the
observed over expected counts.}
\label{fig:aberrant}
\end{figure}

@@

Does the performance of \textit{OneD} for aberrant karyotypes come at the
cost of decreased performance for normal karyotypes? To address this
question, we assembled another data set comprised of mouse B cells and
embryonic stem (ES) cells, both with a normal karyotype The cell types
were pooled in almost equal proportions (see Table~\ref{tab:samples}) and
the same tests as above were performed.

In these conditions, the experimental protocol had a strong effect on the
impact of the different normalization methods (Figure~\ref{fig:normal}A).
For instance, \textit{caICB} and \textit{ICE} increased the similarity
when the protocols were different, but decreased it when the protocols
were the same. The effect was stronger when comparing identical cell
types, but the same trend appeared when comparing different cell types,
indicating that these methods may enhance or reduce biological variation,
depending on the context. Once more, \textit{vanilla} followed the same
trend as \textit{ICE} to a lesser extent. The \textit{LGF} method
increased the similarity when comparing the same cells with different
experimental protocols, and had little to no effect in the other cases.
This indicates that \textit{LGF} is a safe choice in this case.

\textit{OneD} decreased the similarity between different cell types when
using the same protocol and increased it between identical cell types when
using different protocols. In the other two cases, it had little effect.
In summary, \textit{OneD} never enhanced the experimental noise and it
reduced it in one more case than \textit{LGF}.

When interpreting the similarity scores as classification scores, we
observed that all the methods could identify approximately 50\% of the
biological pairs, after which their performance diverged
(Figure~\ref{fig:normal}B). \textit{OneD} achieved the highest area under
the curve on this problem, but with a small margin over the other methods
except \textit{vanilla}.

Using other metrics to compare matrices gave similar results
(Supplementary Figures 6 and 7). \textit{OneD} was always among the top
scoring methods. In these conditions, Spearman correlation of contacts
again appeared as the worst comparison metric because it showed a lower
performance for all the normalization methods.

Taken together, these results indicate that \textit{OneD} performs as well
as the best normalization methods, even with normal karyotypes.

\begin{figure}
\centerline{\includegraphics[width=.50\textwidth]
  {img/correlation_normal_figure4.pdf}}
\caption{
Results of the comparison between samples with normal karyotype.  A.
Average changes compared to raw on the mouse data set. The bars are as in
B. ROC curves on the mouse data set. The legend is as in
Figure~\ref{fig:aberrant}.}
\label{fig:normal}
\end{figure}
